Comparative analysis of PMV Models accuracy implemented in the ISO 7730:2005 and ASHRAE 55:2023

### 2024-09-10 Manuscript/src/sections/appendix.tex
paper: Stefano's comments

In this section, we analyzed the dataset with air speed measurements at three heights.
Only a limited subset of papers in the \ac{db2} had air speed measurements at three heights with \ac{v}~$\geq$~\qty{0.2}{\m\per\s}.
A summary of the papers is provided in Table~\ref{tab:three_heights_papers}.
The table also shows the number of measurements for each paper and how much (percentage) each study contributed to the dataset.
\begin{table}[htb!]
    \centering
    \input{tables/publications_with_three_heights.tex}
    \caption{F1-score for the \ac{pmv} and \ac{pmv-ce} models for different subsets of data.}
    \label{tab:three_heights_papers}
\end{table}
The paper by Cena et al.\ (1999), de Dear et al. (1994), and Zhang et al.\ (2013 and 2010) contributed the most to the dataset.
These \num{3063} entries where used to compare the \ac{pmv} and \ac{pmv-ce} models using the same methodology as in Section~\ref{sec:results}.
The air speed distribution at the three heights for each paper is shown in Figure~\ref{fig:boxenplot_wind_speed_three_heights}.
\begin{figure}[htb!]
    \centering
    \includegraphics[width=\textwidth]{figures/boxenplot_wind_speed_three_heights}
    \caption{Wind speeds distribution at three heights.}
    \label{fig:boxenplot_wind_speed_three_heights}
\end{figure}
The distribution of air speed at the three heights varies between the papers, hence, it is difficult to make generalizations.
However, it should be noted that in most of the studies the air speed measured at the chest height was a good representation of the air speed at the other two heights.
This assumption is not valid for Brenton et al.\ (1994) however this study contributed only \qty{1}{\percent} of the dataset.
The adjustment of \ac{v} based on the \ac{met} significantly increased the base value of the air speed measured by all the studies.
With the value of \ac{v} being \var{delta_v_mean} higher than the average of the other wind speeds combined.
The discussion on weather this increase is justified is beyond the scope of this paper since both Standards require the user to adjust the air speed based on the \ac{met} value.

In this section we are not showing the F1-score for the \ac{pmv} and \ac{pmv-ce} models for the different subsets of data since we have already presented the results in Table~\ref{tab:f1}.
We would like to remind the reader that the F1-macro score for \ac{pmv-ce}~=~\num{.16} was lower than the \ac{pmv}~=~\num{.17} for this subset of data, i.e, \ac{v}~$\geq$~\qty{0.2}{\m\per\s} and \ac{v} measured at three heights.
As previously discussed this is an unexpected result since the \ac{pmv-ce} should be significantly more accurate than the \ac{pmv} model for this subset of data to justify its use.
In this section we are not showing the accuracy of the \ac{pmv} and \ac{pmv-ce} model as a function of the \ac{tsv} since we have a very limited number of measurements for this subset of data moreover this analysis has several limitations as previously discussed.

We, however, plotted the \ac{pmv} and \ac{pmv-ce} values against the \ac{tsv} values and we plotted the \ac{lowess} curves to visualize the prevailing data trend.
The results are shown in Figure~\ref{fig:bubble_models_vs_tsv_three_heights}.
\begin{figure*}[htb!]
    \centering
    \includegraphics[width=\textwidth]{figures/bubble_models_vs_tsv_three_heights}
    \caption{The \ac{lowess} curve shows the relationship between the raw \ac{pmv} and \ac{tsv} data for the dataset with \ac{v}~$\geq$~\qty{0.2}{\m\per\s} and measured at three heights.
    Raw data were then binned and a bubble chart (circle area is proportional to the number of votes in that bin) was superimposed over the regression curve to aid the visualization of a large dataset.
    The brown dashed line represents the identity line where the slope is 1 and the intercept is 0.
    Comparison between the \ac{pmv} and \ac{pmv-ce} models for the dataset with air speed measured at three heights.}
    \label{fig:bubble_models_vs_tsv_three_heights}
\end{figure*}
The results show that the \ac{pmv} model is more accurate than the \ac{pmv-ce} model.
The intercept of the \ac{pmv} model is \num{-.31} while the intercept of the \ac{pmv-ce} model is \num{-.41}.
Moreover, the \ac{pmv-ce} \ac{lowess} curve is only marginally higher than \num{0} despite participants reporting a \ac{tsv} of \num{3}.

Finally, we are showing calculated the overall bias of the \ac{pmv} and \ac{pmv-ce} models.
The results are shown in Figure~\ref{fig:hist_discrepancies_three_heights}.
\begin{figure}[htb!]
    \centering
    \includegraphics[width=\textwidth]{figures/hist_discrepancies_three_heights}
    \caption{Bias of the \ac{pmv} and \ac{pmv-ce} models for the dataset with air speed measured at three heights.}
    \label{fig:hist_discrepancies_three_heights}
\end{figure}
The \ac{pmv} model has a bias of \num{-.28} while the \ac{pmv-ce} model has a bias of \num{-.38}.
The inter-quartile range for the \ac{pmv} model is more centered around \num{0} than the \ac{pmv-ce} model.
These results allow us to conclude that the \ac{pmv} model is more accurate than the \ac{pmv-ce} model for the dataset with air speed measured at three heights.
Consequently, we recommend using the \ac{pmv} model instead of the \ac{pmv-ce}.

%\section{Comparison between the PMV formulations with the \ac{pmvg}}\label{sec:comparison-between-the-pmv-formulations-with-the-two-node-model}
%We compared how the results of the \ac{pmv} and \ac{pmv-ce} perform against the results of the \ac{pmvg}.
%% SS provide more information why this was done
%% Stefano Schiavon: Please provide more details of how these inputs were generated, within which ranges, etc.
%Figure~\ref{fig:pmv_two_node_comparison} was generated by calculating the three \ac{pmv} values using the same combination of \num{10000} randomly generated sets inputs.
%The \ac{pmv} and \ac{pmv-ce} were then plotted on the y-axis while the \ac{pmvg} value was on the x-axis.
%The Figure also shows the \ac{lowess} curves that are used to visualize the prevailing data trend.
%% Stefano Schiavon: Beside the visualization, can you discuss here also the quantitative results to support the conclusion?
%The results of the \ac{pmv} model had a higher agreement than those obtained with the \ac{pmv-ce} model for \ac{pmvg} $\geq$ \qty{0.5}{\m\per\s}.
%% Stefano Schiavon: This is unclear and cannot be deduced from the figure, from where does it come from? How can a PMV model be described by an air speed?
%These results show the \ac{pmv-ce} is less similar to the \ac{pmvg} model than the \ac{pmv}, despite using it in the backend to calculate the cooling effect.
%This is also an unexpected result and we will discuss why this may be in the following paragraph.
%% SS When a figure report how well a variable is predicted, for example x-measured vs x-predicted, the graph must be a square, otherwise it is visually skewed. 1) I suggest to use the words instead of numbers for PMV 2) We should be consistent with the PMV naming we used (do not use ASHRAE and ISO here) 3) why both models are far from Gagge for PMV<0, both model in general produce a cooler feeling than PMVgagge, we should explain why if we keep this paper.
%
%\begin{figure*}[htb!]
%    \centering
%    \includegraphics[width=\textwidth]{figures/pmv_two_node_comparison}
%    \caption{}
%    \label{fig:pmv_two_node_comparison}
%\end{figure*}

>>>>

In this section, we analyzed the dataset with air speed measurements at three heights.
Only a limited subset of papers in the \ac{db2} had air speed measurements at three heights with \ac{v}~$\geq$~\qty{0.2}{\m\per\s}.
A summary of the papers is provided in Table~\ref{tab:three_heights_papers}.
The table also shows the number of measurements for each paper and how much (percentage) each study contributed to the dataset.
\begin{table}[htb!]
    \centering
    \input{tables/publications_with_three_heights.tex}
    \caption{The number of entries used in this analysis grouped by the author/publication.}
    \label{tab:three_heights_papers}
\end{table}
The paper by Cena et al.\ (1999), de Dear et al. (1994), and Zhang et al.\ (2013 and 2010) contributed the most to the dataset (\qty{79}{\percent}).
These \num{3063} entries were used to compare the \ac{pmv} and \ac{pmv-ce} models using the same methodology as in Section~\ref{sec:results}.
The air speed distribution at the three heights for each paper is shown in Figure~\ref{fig:boxenplot_wind_speed_three_heights}.
% todo improvements:
% - it is Bauman, F. not Fred Baumann. The year is missing. 
% -The year is missing for the paper of Benton and Brager. 
% - I would remove the outlier from the graph, the circles. 
% - I would make the figure a bit bigger vertically, it is too compact. 
% -capitalize the Head, Chest, Ankle and Relative
% -order the paper based on their contribution to the dataset
% - call it air speed, not wind speed. 
\begin{figure}[htb!]
    \centering
    \includegraphics[width=\textwidth]{figures/boxenplot_wind_speed_three_heights}
    \caption{Air speeds distribution at three heights.}
    \label{fig:boxenplot_wind_speed_three_heights}
\end{figure}
The distribution of air speed at the three heights varies between the papers, hence, it is difficult to make generalizations.
% todo What do you mean by this? Is the average value close to the chest value? If so, it would be nice to quantify, at least for the overall average of them. 
However, it should be noted that in most of the studies, the air speed measured at the chest height was a good representation of the air speed at the other two heights.
This assumption is not valid for Benton et al.\ (1994); however, this study contributed only \qty{1}{\percent} of the dataset.
The adjustment of \ac{v} based on the \ac{met} significantly increased the base value of the air speed measured by all the studies.
With the value of \ac{v} being \var{delta_v_mean} higher than the average of the other air speeds combined.
The discussion on whether this increase is justified is beyond the scope of this paper, since both standards require the user to adjust the air speed based on the \ac{met} value.

In this section, we are not showing the F1-score for the \ac{pmv} and \ac{pmv-ce} models for the different subsets of data, since we have already presented the results in Table~\ref{tab:f1}.
The F1-macro score for \ac{pmv-ce}~=~\num{.16} was lower than the \ac{pmv}~=~\num{.17} for this subset of data, i.e, \ac{v}~$\geq$~\qty{0.2}{\m\per\s} and \ac{v} measured at three heights.
As previously discussed, this is an unexpected result, since the \ac{pmv-ce} should be more accurate than the \ac{pmv} model for this subset of data to justify its use.
In this Appendix, we are not showing the accuracy of the \ac{pmv} and \ac{pmv-ce} model as a function of the \ac{tsv} since we have too little data for this analysis.

We, however, plotted the \ac{pmv} and \ac{pmv-ce} values against the \ac{tsv} values and we plotted the \ac{lowess} curves to visualize the prevailing data trend.
The results are shown in Figure~\ref{fig:bubble_models_vs_tsv_three_heights}.
\begin{figure*}[htb!]
    \centering
    \includegraphics[width=\textwidth]{figures/bubble_models_vs_tsv_three_heights}
    \caption{Comparison between the \ac{pmv} and \ac{pmv-ce} models for the dataset with air speed measured at three heights. 
    The \ac{lowess} curve shows the relationship between the raw \ac{pmv} and \ac{tsv} data for the dataset with \ac{v}~$\geq$~\qty{0.2}{\m\per\s} and measured at three heights.
    Raw data were then binned and a bubble chart (circle area is proportional to the number of votes in that bin) was superimposed over the regression curve to aid the visualization of a large dataset.
    The brown dashed line represents the identity line, where the slope is 1 and the intercept is 0.}
    \label{fig:bubble_models_vs_tsv_three_heights}
\end{figure*}
The results show that the \ac{pmv} model is more accurate than the \ac{pmv-ce} model.
The intercept of the \ac{pmv} model is \num{-.31} while the intercept of the \ac{pmv-ce} model is \num{-.41}.
Moreover, the \ac{pmv-ce} \ac{lowess} curve is only marginally higher than \num{0} despite participants reporting a \ac{tsv} of \num{3}.

The overall bias of the \ac{pmv} and \ac{pmv-ce} models is shown in Figure~\ref{fig:hist_discrepancies_three_heights}.
\begin{figure}[htb!]
    \centering
    \includegraphics[width=\textwidth]{figures/hist_discrepancies_three_heights}
    \caption{Bias of the \ac{pmv} and \ac{pmv-ce} models for the dataset with air speed measured at three heights.}
    \label{fig:hist_discrepancies_three_heights}
\end{figure}
The \ac{pmv} model has a bias of \num{-.28} while the \ac{pmv-ce} model has a bias of \num{-.38}.
The interquartile range for the \ac{pmv} model is more centered around \num{0} than the \ac{pmv-ce} model.
These results allow us to conclude that the \ac{pmv} model is more accurate than the \ac{pmv-ce} model for the dataset with air speed measured at three heights.

### 2024-09-10 Manuscript/src/sections/conclusions.tex
paper: Stefano's comments

\section{Conclusions}\label{sec:conclusions}
The present study compares the accuracy of the \ac{pmv} model, as used in the \gls{7730} Standard, and its modified version, the \ac{pmv-ce} model, as employed in the \gls{55} Standard.
The \ac{pmv-ce} was specifically developed with the aim of more accurately estimating latent and sensible heat losses from the skin to the environment when air movement is used~\cite{arens_moving_2009}.
We determined the predictive accuracy of the \ac{pmv} and \ac{pmv-ce} models by comparing their results to the \ac{tsv} available in the ASHRAE Global Thermal Comfort Database II v2.1 (Comfort DB).

The overall accuracy of the \ac{pmv} and \ac{pmv-ce} were \qty{33}{\percent} and \qty{34}{\percent}, respectively.
Both models had an accuracy lower than \qty{7}{\percent} when predicting either `warm', `hot', or `cold'.
This value is significantly lower and worse than random guessing, this means that the models are actively misleading the users.
When only comparing the results of the two models for entries with \ac{v}~$\geq$~\qty{0.2}{\m\per\s} -- range in which the two models differ -- the \ac{pmv} outperformed the \ac{pmv-ce} on different metrics.
For example, the median bias of the \ac{pmv} model is \var{bias_median_pmv_0.2} while the median bias for the \ac{pmv-ce} is \var{bias_median_pmv_ce_0.2}.

Our results also show that the classic assumption that people who are `slightly warm' or `slightly cool' are thermally comfortable leads to under-counting thermal discomfort in our dataset.
In the \ac{db2} \qty{68}{\percent} of participants who were `slightly warm' wanted to be `cooler' and \qty{54}{\percent} of participants who were `slightly cool' wanted to be `warmer'.
This finding challenges the above assumption.

\subsection{Implications for the ASHRAE 55 and ISO 7730 Standards}\label{subsec:implications-for-the-ashrae-55-and-iso-7730-standards}
Choosing between the \ac{pmv} and \ac{pmv-ce} is a source of confusion worldwide since both models are widely adopted and incorporated in many building codes and Standards worldwide.
The \gls{55} Standard explains how the \ac{pmv-ce} works, however, we could not find a peer-reviewed publication that clearly explains the benefit of using it.
The \ac{pmv-ce} is more complex, and computationally intensive to run, while not being more accurate than the original \ac{pmv} model.
Hence we do not recommend its use.

The low prediction accuracy of both model formulations, coupled with the lack of data in the extremes of the applicability limits of the models, may suggest that the models' applicability limits should be revisited.
Based on our results we recommend restricting the use of the \ac{pmv} only to values $\lvert \textrm{PMV}\lvert \leq 0.5$.
In other words, we only recommend its use to determine if an environment may be deemed comfortable by a large group of occupants.
The model, however, is not capable of determining the degree by which people are dissatisfied and can only be used to classy if people are on the `cold' (\ac{pmv}~$\leq 0.5$) or `warm' (\ac{pmv}~$\geq 0.5$) side.
Several peer-reviewed papers \cite{Cheung2019, Yao2022, kim2019thermal, tartarini2018thermal, Humphreys2002, doherty_evaluation_1988, tartarini_prediction_2023} provide corroborating evidence that support our results.

We also recommend restricting the ranges on both environmental and personal factors set by both \gls{55} and \gls{7730} Standards.
Currently, these limits extend far beyond the range of data contained in the \ac{db2} dataset.
Being the \ac{db2} the largest thermal comfort dataset in the world and containing data from all continents we deem it not to be necessary to have such wide applicability limits for both \gls{55} and \gls{7730} Standards when these environmental variables are rarely observed in buildings across the world.
In addition, the lack of data available beyond the ranges depicted in Figure~\ref{fig:dist_input_data} did not allow us to test the accuracy of both the \ac{pmv} and \ac{pmv-ce} above those ranges.
Consequently, we suggest reducing both Standards' applicability limits to the ranges shown in Table~\ref{tab:ranges}, until more data are collected.
\begin{table}[htb!]
    \centering
    \begin{tabular}{cc}
        \toprule
        Variable & Proposed range \\
        \midrule
        \ac{tdb} & \qtyrange{17}{30}{\celsius} \\
        \ac{tr} & \qtyrange{17}{30}{\celsius} \\
        \ac{rh} & \qtyrange{10}{80}{\percent} \\
        \ac{clo} & \qtyrange{0.3}{1.5}{clo} \\
        \ac{met} & \qtyrange{1}{2}{met} \\
        \bottomrule
    \end{tabular}
    \caption{New proposed applicability limits for the \gls{55} and \gls{7730} Standards.}
    \label{tab:ranges}
\end{table}
\mycite{Humphreys2002} reached a similar conclusion and observed that the ranges of inputs that are consistent with a valid use of PMV are much narrower than those given in \gls{7730}.
The limit on \ac{met} is also corroborated by evidence provided by \mycite{doherty_evaluation_1988}.

The \ac{tsv} is ambiguous and does not allow determining if an action should be taken to increase the participant's thermal comfort.
Assuming that people who are `slightly warm' or `slightly cool' are comfortable leads to undercounting thermal discomfort in our dataset, and it is not a valid assumption.
We recommend the use of the thermal preference scale to determine how people perceive their thermal environment, this scale more clearly allows an HVAC system to decide whether to change or not the thermal environment.

>>>>

\section{Conclusions}\label{sec:conclusions}
The present study compares the accuracy of the \ac{pmv} model, as used in the \gls{7730} standard, and its modified version, the \ac{pmv-ce} model, as employed in the \gls{55} standard.
The \ac{pmv-ce} was intentionally developed to more accurately estimate the effect of elevated air movement~\cite{arens_moving_2009}.
We determined the predictive accuracy of the \ac{pmv} and \ac{pmv-ce} models by comparing their results to the \ac{tsv} available in the ASHRAE Global Thermal Comfort Database II v2.1.

% todo  could you add the values of F1-macro here as well?
The overall accuracy of the \ac{pmv} and \ac{pmv-ce} were \qty{33}{\percent} and \qty{34}{\percent}, respectively.
Both models had an accuracy lower than \qty{7}{\percent} when predicting either `warm', `hot', or `cold'.
This value is lower than random guessing (\qty{14}{\percent}), this means that the models are misleading the users.
When only comparing the results of the two models for entries with \ac{v}~$\geq$~\qty{0.2}{\m\per\s} the \ac{pmv} outperformed the \ac{pmv-ce} on different metrics.
For example, the median bias of the \ac{pmv} model is \var{bias_median_pmv_0.2} while the median bias for the \ac{pmv-ce} is \var{bias_median_pmv_ce_0.2}.

Our results also show that the assumption that people who are `slightly warm' or `slightly cool' are thermally comfortable is incorrect and leads to under-counting thermal discomfort.
In the \ac{db2} \qty{68}{\percent} of participants who were `slightly warm' wanted to be `cooler' and \qty{54}{\percent} of participants who were `slightly cool' wanted to be `warmer'.

\subsection{Implications for the ASHRAE 55 and ISO 7730 standards}\label{subsec:implications-for-the-ashrae-55-and-iso-7730-standards}
As mentioned previously, researchers, educators and practitioners worldwide may wonder which model they should use.
The \gls{55} standard explains how the \ac{pmv-ce} works, however, we could not find a peer-reviewed publication that proves it has superior prediction performances.
The \ac{pmv-ce} is more complex, and computationally intensive to run, while not being more accurate than the original \ac{pmv} model.
Hence, we do not recommend its use.

The very low prediction accuracy of both model formulations, along with the lack of data at the extremes of their applicability limits, suggests that these limits may need to be reconsidered.
Based on our results, we recommend restricting the use of the \ac{pmv} only to values $\lvert \textrm{PMV}\lvert \leq 0.5$.
In other words, we only recommend it for determining whether an environment may be deemed comfortable by a large group of occupants.
The model, however, is not capable of determining the degree by which people are dissatisfied and can only be used to classify if individuals are on the `cold' (\ac{pmv}~$\leq 0.5$) or `warm' (\ac{pmv}~$\geq 0.5$) side.
Several peer-reviewed papers \cite{Cheung2019, Yao2022, kim2019thermal, tartarini2018thermal, Humphreys2002, doherty_evaluation_1988} provide corroborating evidence that supports our results.

We also recommend restricting the ranges on both environmental and personal factors set by both \gls{55} and \gls{7730} standards.
Currently, these limits extend far beyond the range of data contained in the \ac{db2} dataset.
Being the \ac{db2} the largest thermal comfort dataset in the world and containing data from all continents, we deem it not to be necessary to have such wide applicability limits for both \gls{55} and \gls{7730} standards when these environmental variables are rarely observed in buildings across the world.
In addition, the lack of data available beyond the ranges depicted in Figure~\ref{fig:dist_input_data} did not allow us to test the accuracy of both the \ac{pmv} and \ac{pmv-ce} above those ranges.
Consequently, we suggest reducing both standards' applicability limits to the ranges shown in Table~\ref{tab:ranges}, until more data are collected.
\begin{table}[htb!]
    \centering
    \begin{tabular}{cc}
        \toprule
        Variable & Proposed range \\
        \midrule
        \ac{tdb} & \qtyrange{17}{30}{\celsius} \\
        \ac{tr} & \qtyrange{17}{30}{\celsius} \\
        \ac{rh} & \qtyrange{10}{80}{\percent} \\
        \ac{clo} & \qtyrange{0.3}{1.5}{clo} \\
        \ac{met} & \qtyrange{1}{2}{met} \\
        \bottomrule
    \end{tabular}
    \caption{New proposed applicability limits for the \gls{55} and \gls{7730} standards.}
    \label{tab:ranges}
\end{table}
\mycite{Humphreys2002} reached a similar conclusion and observed that the ranges of inputs consistent with a valid use of PMV are much narrower than those given in \gls{7730}.
The limit on \ac{met} is also corroborated by evidence provided by \mycite{doherty_evaluation_1988}.

The \ac{tsv} is ambiguous and does not allow determining if an action should be taken to increase the participant's thermal comfort.
Assuming that people who are `slightly warm' or `slightly cool' are comfortable leads to undercounting thermal discomfort in our dataset, which is not a valid assumption.
We recommend the use of the thermal preference scale to determine how people perceive their thermal environment. 
This scale more clearly allows an HVAC system to decide whether to change or not the thermal environment.
